Methods for Analyzing Models of\\ Rod-Shaped Bacteria

### 2025-09-29 paper/main.tex
update todos

Despite their popularity, \acp{abm} face multiple challenges in practise.
Constructing a fully working \ac{abm} simulation from scratch is an intricate process that requires
a large set of skills and time.
It is therefore common practise to reuse existing simulation tools, which brings along a number of
other problems.
Even if the existing tool is designed to be generalizable, it often comes with a pre-defined set
of parameters that need to be specified.
Furthermore, as we have discussed in our previous work~\cite{Pleyer2023}, these tools were most
often not designed to be flexible but rather built for specific purposes thus limiting their
generalizability.
% TODO citations
Typically researchers aim to estimate every individual parameter of their model.
This is achieved using methods such as the profile-likelihood and model reduction techniques.
% TODO citaitons
When it comes to \acp{abm} however, it is often only shown that the results which have been
generated by an \ac{abm} lead to the desired phenomenon which aligns with the observed biological
reality.
But this process does not actually estimate the necessary parameters directly
% TODO citations
and reduces the confidence in the constructed models and raises concerns with respect to the
validity of the obtained results.

@@

%###################################################################################################
\section{Application 1: Elastic deformations of \textit{E.Coli} (TODO)}
\label{section:application-1}

@@

We can again calculate likelihood profiles for every optimized parameter.
The results displayed in Figure~\ref{fig:likelihood-profiles-comparison-with-division} show that
some curves exhibit a plateau close to the calculated optimum of the estimated parameter.
This concerns the division lengths and some growth rates of the mother bacteria which both determine
if and when a cell divides.
Such behaviour can be attributed to the missing images around the division events of the cells which
were not able to be segmented.
\textbf{TODO check this with data}
It can be argued that this leads to a non-identifiability for the first division length while the
remaining parameters are still identifiable although the effect is also notable.
These results highlight a reassuring finding, namely that the method of profile-likelihood is able
to properly display this lack of data.
It therefore further underlines the accuracy and predictive power of our methods.

>>>>

Despite their popularity, \acp{abm} face multiple challenges in practise.
Constructing a fully working \ac{abm} simulation from scratch is an intricate process that requires
a large set of skills and time.
It is therefore common practise to reuse existing simulation tools, which brings along a number of
other problems.
Even if the existing tool is designed to be generalizable, it often comes with a pre-defined set
of parameters that need to be specified.
Furthermore, as we have discussed in our previous work~\cite{Pleyer2023}, these tools were most
often not designed to be flexible but rather built for specific purposes thus limiting their
generalizability \textbf{TODO CITATION}.
Typically researchers aim to estimate every individual parameter of their model.
This is achieved using methods such as the profile-likelihood and model reduction
techniques \textbf{TODO CITATION}.
When it comes to \acp{abm} however, it is often only shown that the results which have been
generated by an \ac{abm} lead to the desired phenomenon which aligns with the observed biological
reality.
But this process does not actually estimate the necessary parameters directly \textbf{TODO CITATION}
and reduces the confidence in the constructed models and raises concerns with respect to the
validity of the obtained results.

@@

%###################################################################################################
\section{Application 1: Elastic deformations of \textit{E.Coli}}
\label{section:application-1}

@@

We can again calculate likelihood profiles for every optimized parameter.
The results displayed in Figure~\ref{fig:likelihood-profiles-comparison-with-division} show that
some curves exhibit a plateau close to the calculated optimum of the estimated parameter.
This concerns the division lengths and some growth rates of the mother bacteria which both determine
if and when a cell divides.
Such behaviour can be attributed to the missing images around the division events of the cells which
were not able to be segmented.
\textbf{TODO check this with results}
It can be argued that this leads to a non-identifiability for the first division length while the
remaining parameters are still identifiable although the effect is also notable.
These results highlight a reassuring finding, namely that the method of profile-likelihood is able
to properly display this lack of data.
It therefore further underlines the accuracy and predictive power of our methods.
